---
title: "BINF6310_ Final_Project"
author: "Bobby Luker"
date: "2026-04-27"
output: pdf_document
---

# Initialize my environment and establish all needed libraries
```{r, message=FALSE, warning=FALSE}
rm(list=ls())
setwd("/Users/bobbyluker/Desktop/UNCC/BINF6310/Project/data/")

library(readxl) # Reads in excel files

library(stringr) # Cleans column names
library(dplyr) # Data wrangling

library(limma) # Used for statistical modeling

library(ggplot2) # Data visualization
library(ggrepel) # Volcano plot options

library(clusterProfiler) # Used for GO enrichment and GSEA
library(org.Hs.eg.db) # Maps genes to EntrezIDs for GO enrichment
library(enrichplot) # GO enrichment plotting

library(knitr) # Knits markdown tables to pdf
library(kableExtra) # Formats tables
```


# Import the data file and prep for PCA
```{r}
lung_data <- read.csv("GSE40419_LC-87_RPKM_expression.csv")

# Extract expression values only
expr_data <- lung_data[, 7:ncol(lung_data)]

# Add gene column for aggregation
expr_data$gene <- lung_data$gene

# Aggregate duplicate genes by mean expression
expr_data <- aggregate(. ~ gene, data = expr_data, FUN = mean)

# Set row names and remove gene column
rownames(expr_data) <- expr_data$gene
expr_data$gene <- NULL

# Filter low-expression genes
# filtered_lung_data <- expr_data[rowMeans(expr_data) > 0.5, ]
filtered_lung_data <- expr_data[rowSums(expr_data > 1) >= (0.2 * ncol(expr_data)),]

# Log transform with a pseudocount of 1
log_fld <- log2(filtered_lung_data + 1)

# Transpose for PCA
trans_fld <- t(log_fld)
```


# Import the metadata
```{r}
metadata <- read_excel("SuppTable1.xlsx")

# Clean column names (remove spaces and line breaks)
colnames(metadata) <- str_trim(str_replace_all(colnames(metadata), "[\r\n]", ""))

# Keep only relevant columns
metadata <- metadata %>% dplyr::select(
  Publication_index,
  Smoking_status = `Smoking_status(0=neverSmoker;1=smoker;2=current_smoker;3=unknown)`,
  Cancer_RNAseq = `Cancer_RNAseq(0=no;1=yes)`,
  Normal_RNAseq = `Normal_RNAseq(0=no;1=yes)`
)

# Define Smoking_Group
metadata$Smoking_Group <- case_when(
  metadata$`Smoking_status` %in% c(1,2) ~ "Smoker",
  metadata$`Smoking_status` == 0 ~ "Non_Smoker",
  TRUE ~ "Unknown"
)

# Categorize the "_nor" samples as smoker, non-smoker, or unknown
sample_ids <- rownames(trans_fld)
base_ids <- sub("_nor$", "", sample_ids)
metadata <- metadata[match(base_ids, metadata$Publication_index), ]
```



# Creating table containing the sample size by tissue and smoker status
```{r}
# Establishing the tumor and non-tumor group sizes
tissue <- c(rep("Tumor", 87), rep("Normal", 77))

# Creating data frame of sample sizes by tissue and smoker status
sample_count_df <- data.frame(
  Group = tissue,
  Smoking = metadata$Smoking_Group
)

# Converting data frame into a table
sample_count_table <- sample_count_df %>%
  group_by(Group, Smoking) %>%
  summarise(n = n(), .groups = "drop") %>%
  tidyr::pivot_wider(names_from = Smoking, values_from = n, values_fill = 0)

# Preparing table to be knit to PDF
sample_count_table %>%
  kable(format = "latex", caption = "Sample Sizes by Group and Smoking Status") %>%
  kable_styling(full_width = FALSE, position = "center") %>%
  column_spec(1, bold = TRUE)
```

# p-value diagnostics function
```{r}
plot_limma_diagnostics <- function(fit, coef_name, title_prefix = "") {
  
  pvals <- fit$p.value[, coef_name]
  logFC <- fit$coefficients[, coef_name]
  
  par(mfrow = c(1,2))
  
  # 1. P-value histogram
  hist(pvals,
       breaks = 50,
       main = paste(title_prefix, "P-value histogram"),
       xlab = "p-values",
       col = "grey")
  
  # 2. QQ-style p-value plot (your professor’s Figure 4 style)
  p_sorted <- sort(pvals)
  expected <- (1:length(p_sorted)) / length(p_sorted)
  
  plot(expected, p_sorted,
       main = paste(title_prefix, "P-value plot (QQ style)"),
       xlab = "Expected (Uniform)",
       ylab = "Observed p-values")
  
  abline(0,1,col="red")
  
  par(mfrow = c(1,1))
}
```

# Paired Smoker Tissue Differential Expression Analysis using "limma" 
```{r}
# Removing '_nor' from normal samples to correctly pair data
sample_names <- colnames(log_fld)
sample_clean <- sub("_nor$", "", sample_names)

# Defining patient and grouping
patient <- factor(sample_clean)
group <- ifelse(grepl("_nor$", sample_names), "Normal", "Tumor")
group <- factor(group, levels = c("Normal", "Tumor"))

# Checking that there are 77 pairs.
metadata_paired <- table(patient, group)

metadata_paired <- rownames(metadata_paired)[
  metadata_paired[, "Normal"] > 0 &
  metadata_paired[, "Tumor"] > 0
]

# Selecting the patients that have tumor and normal tissue samples
keep <- patient %in% metadata_paired

log_fld_paired <- log_fld[, keep]
patient <- droplevels(patient[keep])
group <- droplevels(group[keep])

# Code to verify that there are 77 patient pairs
# cat("Patients with incomplete pairs (MUST be empty):\n")
# print(table(patient)[table(patient) != 2])
# 
# cat("\nFinal sample counts:\n")
# print(table(group))
# cat("\nTotal patients:\n")
# print(length(unique(patient)))

# Create matrix for limma fitting
paired_matrix <- model.matrix(~ patient + group)

# Fit linear model using limma, eBayes stabilizes variance estimates
fit <- lmFit(log_fld_paired, paired_matrix)
fit <- eBayes(fit, trend=TRUE)

plot_limma_diagnostics(fit, "groupTumor", "PAIRED Tumor vs Normal")

# Extract result
results_paired <- topTable(
  fit,
  coef = "groupTumor",
  number = Inf,
  adjust.method = "BH"
)

results_paired$gene <- rownames(results_paired)

# Label Significance of DEGs
results_paired$Regulation <- "Not Significant"
results_paired$Regulation[
  results_paired$adj.P.Val < 0.05 & results_paired$logFC > 1
] <- "Upregulated"
results_paired$Regulation[
  results_paired$adj.P.Val < 0.05 & results_paired$logFC < -1
] <- "Downregulated"

# Collecting the top 10 most significant DEGs for the plot
top_genes <- results_paired %>%
  dplyr::filter(adj.P.Val < 0.01 & abs(logFC) > 2) %>%
  dplyr::arrange(adj.P.Val) %>%
  head(10)

# Counts of DEGs
results_paired %>%
  filter(adj.P.Val < 0.05 & abs(logFC) > 1) %>%
  summarise(
    Total = n(),
    Upregulated = sum(logFC > 1),
    Downregulated = sum(logFC < -1)
  )

# Create volcano plot
volcano_paired <- ggplot(results_paired, aes(x = logFC, y = -log10(adj.P.Val))) +
  
  geom_point(aes(color = Regulation), alpha = 0.6, size = 1.5) +
  
  #Highlight and label top 10 DEGs
  geom_point(
    data = top_genes,
    aes(x = logFC, y = -log10(adj.P.Val)),
    color = "orange",
    size = 2
  ) +
  
  geom_text_repel(
    data = top_genes,
    aes(label = gene),
    size = 3,
    max.overlaps = 20
  ) +
  
  # Threshold lines
  geom_vline(xintercept = c(-1, 1), linetype = "dashed") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  
  # Point colors
  scale_color_manual(values = c(
    "Upregulated" = "red",
    "Downregulated" = "blue",
    "Not Significant" = "grey"
  )) +
  
  # Labels
  theme_minimal() +
  labs(
    x = "Log2 Fold Change",
    y = "-log10 Adjusted P-value",
    color = "Regulation"
  )

print(volcano_paired)
```

# Paired DEG GO Enrichment
```{r}
# Define paired DEGs from the results_paired
paired_deg <- results_paired %>%
  dplyr::filter(adj.P.Val < 0.05 & abs(logFC) > 1)

# Convert gene symbols to Entrez IDs
paired_deg_ids <- bitr(
  paired_deg$gene,
  fromType = "SYMBOL",
  toType = "ENTREZID",
  OrgDb = org.Hs.eg.db
) %>%
  dplyr::distinct(ENTREZID, .keep_all = TRUE)

# Remove duplicates (safety step)
paired_deg_ids <- paired_deg_ids[!duplicated(paired_deg_ids$ENTREZID), ]

# Define background by all genes
background_ids <- bitr(
  rownames(results_paired),
  fromType = "SYMBOL",
  toType = "ENTREZID",
  OrgDb = org.Hs.eg.db
) %>%
  dplyr::distinct(ENTREZID, .keep_all = TRUE)

background_ids <- background_ids[!duplicated(background_ids$ENTREZID), ]

# GO enrichment
go_paired <- enrichGO(
  gene = paired_deg_ids$ENTREZID,
  universe = background_ids$ENTREZID,
  OrgDb = org.Hs.eg.db,
  ont = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff = 0.05,
  readable = TRUE
)

# Create Plot
go_enrich_paired_DEGs <- dotplot(go_paired, showCategory = 12) +
  theme_minimal(base_size = 12) +
  scale_color_continuous(
    low = "blue",
    high = "red",
    name = "Adj p-value"
  )

print(go_enrich_paired_DEGs)
```

# Paired Smoker Tissue interaction with Smoking Status Differential Expression Analysis using "limma" 
```{r}
# Use paired dataset (already created earlier: log_fld_paired)
sample_names <- colnames(log_fld_paired)

# Create metadata_interaction 
metadata_interaction <- metadata[match(patient, metadata$Publication_index), ]
metadata_interaction$Smoking_Group <- factor(metadata_interaction$Smoking_Group)
metadata_interaction$Tissue <- group
metadata_interaction$Patient <- patient

# Model matrix
interaction_matrix <- model.matrix(~ Patient + Tissue * Smoking_Group, data = metadata_interaction)

# Fit model
interaction_fit <- lmFit(log_fld_paired, interaction_matrix)
interaction_fit <- eBayes(interaction_fit, trend=TRUE)

# Extract interaction term
interaction_results <- topTable(
  interaction_fit,
  coef = "TissueTumor:Smoking_GroupSmoker",
  number = Inf,
  adjust.method = "BH"
)

interaction_results$gene <- rownames(interaction_results)

# Significance labeling
interaction_results$Significance <- "Not Significant"
interaction_results$Significance[
  interaction_results$adj.P.Val < 0.05 & interaction_results$logFC > 0
] <- "Positive Interaction"
interaction_results$Significance[
  interaction_results$adj.P.Val < 0.05 & interaction_results$logFC < 0
] <- "Negative Interaction"

# Top genes (same logic, unchanged)
top_genes <- interaction_results %>%
  dplyr::filter(adj.P.Val < 0.05 & abs(logFC) > 1) %>%
  dplyr::arrange(adj.P.Val) %>%
  head(10)

# DEG summary (same logic, unchanged)
interaction_results %>%
  dplyr::filter(adj.P.Val < 0.05 & abs(logFC) > 0) %>%
  dplyr::summarise(
    Total = n(),
    Upregulated = sum(logFC > 0),
    Downregulated = sum(logFC < 0)
  )

# Volcano plot (unchanged structure)
volcano_interaction <- ggplot(interaction_results,
       aes(x = logFC, y = -log10(adj.P.Val))) +
  
  geom_point(aes(color = Significance), alpha = 0.6, size = 1.5) +
  
  geom_point(data = top_genes,
             aes(x = logFC, y = -log10(adj.P.Val)),
             color = "orange", size = 2.2) +
  
  geom_text_repel(data = top_genes,
                  aes(label = gene),
                  size = 3,
                  max.overlaps = 20,
                  box.padding = 0.4,
                  point.padding = 0.3,
                  segment.color = "grey40") +
  
  geom_vline(xintercept = 0, linetype = "dashed") +
  
  scale_color_manual(values = c(
    "Positive Interaction" = "red",
    "Negative Interaction" = "blue",
    "Not Significant" = "grey"
  )) +
  
  theme_minimal() +
  
  labs(
    x = "Interaction Effect (Smoking x Tumor vs Normal)",
    y = "-log10 Adjusted P-value",
    color = "Significance"
  )

print(volcano_interaction)
```

# GSEA plot of the interaction results
```{r}
gene_list <- interaction_results$logFC
names(gene_list) <- interaction_results$gene

gsea <- gseGO(
  geneList = sort(gene_list, decreasing = TRUE),
  OrgDb = org.Hs.eg.db,
  ont = "BP",
  keyType = "SYMBOL"
)

gsea_plot <- dotplot(gsea, showCategory = 12, color = "NES") +
  theme_minimal(base_size = 12) + 
  scale_color_gradient2(
    low = "blue",
    mid = "white",
    high = "red",
    midpoint = 0
  )

 print(gsea_plot)
```


# PCA Analysis
```{r}
# Executing PCA
pca <- prcomp(trans_fld, scale. = TRUE)

# Buiding PCA data frame for plotting
pca_df <- data.frame(
  Sample = rownames(pca$x),
  PC1 = pca$x[,1],
  PC2 = pca$x[,2],
  Group = tissue,
  Smoking_Group = metadata$Smoking_Group
)

# Creating the PCA plot
pca_plot <- ggplot(pca_df, aes(x = PC1, y = PC2, color = Group, fill = Group, shape = Smoking_Group)) +
  geom_point(size = 2.5, stroke = 0.5, color = "black") +
  
  # Ellipses only for smoker and non-smoker
  stat_ellipse(
    data = subset(pca_df, Smoking_Group %in% c("Smoker", "Non_Smoker")),
    aes(linetype = Smoking_Group),
    type = "norm", linewidth = 1, alpha = 0.6
  ) +
  
  theme_minimal() +
  scale_color_manual(values = c("Tumor" = "salmon", "Normal" = "#4DB6AC")) +
  scale_fill_manual(values = c("Tumor" = "salmon", "Normal" = "#4DB6AC")) +
  scale_shape_manual(values = c("Non_Smoker" = 21, "Smoker" = 24, "Unknown" = 22)) +
  labs(
    x = paste0("PC1 (", round(summary(pca)$importance[2,1]*100,1), "% variance)"),
    y = paste0("PC2 (", round(summary(pca)$importance[2,2]*100,1), "% variance)")
  )

print(pca_plot)
```

# Unpaired Tissue Differential Expression Analysis using "limma" 
```{r}
# Create matrix for limma fitting
unpaired_matrix <- model.matrix(~ tissue)
colnames(unpaired_matrix) <- c("Intercept", "Tumor_vs_Normal")

# Fit linear model using limma, eBayes stabilizes variance estimates
unpaired_fit <- lmFit(log_fld, unpaired_matrix)
unpaired_fit <- eBayes(unpaired_fit, trend=TRUE)

plot_limma_diagnostics(unpaired_fit, "Tumor_vs_Normal", "UNPAIRED Tumor vs Normal")

# Extract result
unpaired_results <- topTable(unpaired_fit,
                    coef = "Tumor_vs_Normal",
                    number = Inf,
                    adjust.method = "BH")

# Labeling significance of DEGs
unpaired_results$gene <- rownames(unpaired_results)
unpaired_results$Regulation <- "Not Significant"
unpaired_results$Regulation[unpaired_results$adj.P.Val < 0.05 & unpaired_results$logFC > 1] <- "Upregulated"
unpaired_results$Regulation[unpaired_results$adj.P.Val < 0.05 & unpaired_results$logFC < -1] <- "Downregulated"

# Collecting the top 10 most significant DEGs for the plot
top_genes <- unpaired_results %>%
  dplyr::filter(adj.P.Val < 0.01 & abs(logFC) > 2) %>%
  dplyr::arrange(adj.P.Val) %>%
  head(10)

# Counts of DEGs
unpaired_results %>%
  filter(adj.P.Val < 0.05 & abs(logFC) > 1) %>%
  summarise(
    Total = n(),
    Upregulated = sum(logFC > 1),
    Downregulated = sum(logFC < -1)
  )

# Creating volcano plot
volvano_tumor_v_normal <- ggplot(unpaired_results, 
  aes(x = logFC, y = -log10(adj.P.Val))) +
  geom_point(aes(color = Regulation), alpha = 0.6) +
  
  # Highlight and label top 10 DEGs
  geom_point(data = top_genes, aes(x = logFC, y = -log10(adj.P.Val)), color = "orange", size = 2) +
  geom_text_repel(data = top_genes,
                  aes(label = gene),
                  size = 3,
                  max.overlaps = 20) +
  
  # Threshold lines
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "black") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black") +
  
  # Point colors
  scale_color_manual(values = c(
    "Upregulated" = "red",
    "Downregulated" = "blue",
    "Not Significant" = "grey"
  )) +
  
  # Labels & theme
  labs(
       x = "Log2 Fold Change (Tumor vs Normal)",
       y = "-log10 Adjusted P-value",
       color = "Regulation") +
  theme_minimal()

print(volvano_tumor_v_normal)
```

# Unpaired Non-Smoker Tissue Differential Expression Analysis using "limma" 
```{r}
# Create combined factor of tumor/normal and smoking status
Smoking_Groups <- factor(paste(tissue, metadata$Smoking_Group, sep = "_"))

# Create matrix for limma fitting. Setting intercept to 0 to explicitly compare the smoking conditions. Otherwise, the non_smoker group will be compared to a hidden baseline group mean (intercept)
smoker_group_matrix <- model.matrix(~0 + Smoking_Groups)
colnames(smoker_group_matrix) <- make.names(levels(Smoking_Groups)) # ensures valid names

# Fit linear model
non_smoker_fit <- lmFit(log_fld, smoker_group_matrix)

# Defining a matrix to compare specific groups
nonsmoker_contrast_matrix <- makeContrasts(
  Tumor_Non_Smoker_vs_Normal_Non_Smoker = Tumor_Non_Smoker - Normal_Non_Smoker,
  levels = smoker_group_matrix
)

# Fit linear model using limma, eBayes stabilizes variance estimates
non_smoker_fit <- contrasts.fit(non_smoker_fit, nonsmoker_contrast_matrix)
non_smoker_fit <- eBayes(non_smoker_fit, trend=TRUE)

plot_limma_diagnostics(non_smoker_fit,
                       "Tumor_Non_Smoker_vs_Normal_Non_Smoker",
                       "NON-SMOKER contrast")

# Extract result
results_non_smoker <- topTable(non_smoker_fit, number = Inf)
results_non_smoker$gene <- rownames(results_non_smoker)

# Labeling Significance of DEGs
results_non_smoker$Significance <- "Not Significant"
results_non_smoker$Significance[results_non_smoker$adj.P.Val < 0.05 & results_non_smoker$logFC > 1] <- "Upregulated"
results_non_smoker$Significance[results_non_smoker$adj.P.Val < 0.05 & results_non_smoker$logFC < -1] <- "Downregulated"

# Collecting the top 10 most significant DEGs for the plot
top_genes <- results_non_smoker %>%
  dplyr::filter(adj.P.Val < 0.01 & abs(logFC) > 2) %>%
  head(10)

# Counts of DEGs
results_non_smoker %>%
  filter(adj.P.Val < 0.05 & abs(logFC) > 1) %>%
  summarise(
    Total = n(),
    Upregulated = sum(logFC > 1),
    Downregulated = sum(logFC < -1)
  )

# Creating volcano plot
volcano_non_smoker <- ggplot(results_non_smoker, aes(x = logFC, y = -log10(adj.P.Val))) +
  geom_point(aes(color = Significance), alpha = 0.6, size = 1.5) +
  
  # Highlight and label top 10 DEGs
  geom_point(data = top_genes, aes(x = logFC, y = -log10(adj.P.Val)), color = "orange", size = 2) +
  geom_text_repel(data = top_genes, aes(label = gene), size = 3, max.overlaps = 20) +
  
  # Threshold lines
  geom_vline(xintercept = c(-1, 1), linetype = "dashed") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  
  # Point colors
  scale_color_manual(values = c(
    "Upregulated" = "red",
    "Downregulated" = "blue",
    "Not Significant" = "grey"
  )) +
  
  theme_minimal() +
  theme(legend.key.size = unit(0.5, "cm")) +
  labs(
       x = "Log2 Fold Change (Non-Smoker: Tumor vs Normal)",
       y = "-log10 Adjusted P-value",
       color = "Regulation")

print(volcano_non_smoker)
```

# Unpaired Smoker Tissue Differential Expression Analysis using "limma" 
```{r}
# Fit linear model
smoker_fit <- lmFit(log_fld, smoker_group_matrix)

# Create matrix for limma fitting
smoker_contrast_matrix <- makeContrasts(
  Tumor_Smoker_vs_Normal_Smoker = Tumor_Smoker - Normal_Smoker,
  levels = smoker_group_matrix
)

# Fit linear model using limma, eBayes stabilizes variance estimates
smoker_fit <- contrasts.fit(smoker_fit, smoker_contrast_matrix)
smoker_fit <- eBayes(smoker_fit, trend=TRUE)

plot_limma_diagnostics(smoker_fit,
                       "Tumor_Smoker_vs_Normal_Smoker",
                       "SMOKER contrast")

# Extract result
results_smoker <- topTable(smoker_fit, number = Inf)
results_smoker$gene <- rownames(results_smoker)

# Labeling Significance of DEGs
results_smoker$Significance <- "Not Significant"
results_smoker$Significance[results_smoker$adj.P.Val < 0.05 & results_smoker$logFC > 1] <- "Upregulated"
results_smoker$Significance[results_smoker$adj.P.Val < 0.05 & results_smoker$logFC < -1] <- "Downregulated"

# Collecting the top 10 most significant DEGs for the plot
top_genes <- results_smoker %>%
  dplyr::filter(adj.P.Val < 0.01 & abs(logFC) > 2) %>%
  head(10)

# Counts of DEGs
results_smoker %>%
  filter(adj.P.Val < 0.05 & abs(logFC) > 1) %>%
  summarise(
    Total = n(),
    Upregulated = sum(logFC > 1),
    Downregulated = sum(logFC < -1)
  )

# Creating Volcano plot
volcano_smoker <- ggplot(results_smoker, aes(x = logFC, y = -log10(adj.P.Val))) +
  geom_point(aes(color = Significance), alpha = 0.6, size = 1.5) +
  
  # Highlight and label top 10 DEGs
  geom_point(data = top_genes, aes(x = logFC, y = -log10(adj.P.Val)), color = "orange", size = 2) +
  geom_text_repel(data = top_genes, aes(label = gene), size = 3, max.overlaps = 20) +
  
  # Threshold lines
  geom_vline(xintercept = c(-1, 1), linetype = "dashed") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  
  # Point Colors
  scale_color_manual(values = c(
    "Upregulated" = "red",
    "Downregulated" = "blue",
    "Not Significant" = "grey"
  )) +
  
  # Labels and theme
  theme_minimal() +
  theme(legend.key.size = unit(0.5, "cm")) +
  labs(
       x = "Log2 Fold Change (Smoker: Tumor vs Normal)",
       y = "-log10 Adjusted P-value",
       color = "Regulation")

print(volcano_smoker)
```



# Table of the 3 most differentiated genes for each group
```{r}
# Function to select the top 3 up/down genes and bind the rows together
get_top_genes <- function(df, top_n = 3, logfc_cut = 1, pval_cut = 0.05, group_name = "") {
  up <- df %>%
    filter(logFC > logfc_cut & adj.P.Val < pval_cut) %>%
    arrange(adj.P.Val) %>%
    head(top_n) %>%
    mutate(Regulation = "Upregulated", Comparison = group_name) %>%
    dplyr::select(Comparison, Regulation, gene, logFC, adj.P.Val)
  
  down <- df %>%
    filter(logFC < -logfc_cut & adj.P.Val < pval_cut) %>%
    arrange(adj.P.Val) %>%
    head(top_n) %>%
    mutate(Regulation = "Downregulated", Comparison = group_name) %>%
    dplyr::select(Comparison, Regulation, gene, logFC, adj.P.Val)
  
  rbind(up, down)
}

# Create summary_table from the DE results
summary_table <- rbind(
  get_top_genes(unpaired_results, group_name = "Tumor vs Normal (All)"),
  get_top_genes(results_non_smoker, group_name = "Tumor vs Normal (Non-smokers)"),
  get_top_genes(results_smoker, group_name = "Tumor vs Normal (Smokers)"),
  get_top_genes(results_paired, group_name = "Paired"),
  get_top_genes(interaction_results, group_name = "Paired Smoker Interaction")
)

# dentify shared genes across groups (Up/Down separately)
shared_genes <- summary_table %>%
  group_by(Regulation, gene) %>%
  summarize(n_groups = n_distinct(Comparison), .groups = "drop") %>%
  filter(n_groups > 1) %>%
  pull(gene)

# Add asterisk for shared genes
summary_table_marked <- summary_table %>%
  mutate(gene = ifelse(gene %in% shared_genes, paste0(gene, "*"), gene)) %>%
  dplyr::select(-adj.P.Val)   # remove adjusted p-value column

# Remove row names
rownames(summary_table_marked) <- NULL
```
\newpage
```{r}
# Print clean table
kable(summary_table_marked, caption = "Top upregulated and downregulated genes identified in tumor vs normal analyses across all samples, paired samples, smokers, non-smokers, and smoking interaction analysis (* indicates genes significant in multiple comparisons)") %>%
  kable_styling(full_width = FALSE, position = "center", font_size = 10)
```


